Testing the Relationship between Phenomenological Control related to Illusion Sensitivity

### 2024-04-26 paper/manuscript.qmd
small changes to writing
Table

  - name: Zoltan Dienes
    orcid: 0000-0001-7454-3161

@@

Visual illusions highlight how easily our conscious experience can be altered with respect to perceptual reality. Despite sharing in-principle mechanisms with phenomenological control, i.e., the ability to alter our perceptual experience to match task demands or expectations, research tying the two remains scarce. This study aims to replicate and expand Lush et al. (2022) reporting an absence of correlation between phenomenological control (measured using the Phenomenological Control Scale) and illusion sensitivity to different illusion types. *[N participants were recruited in an online study. Results will added in the final version of the manuscript]*.

@@

  references-meta-analysis: "References marked with an asterisk indicate studies included in the meta-analysis."

@@

Interestingly, @lush2022 did test for a relationship between PC and illusion sensitivity using the Müller-Lyer illusion (in which the arrangement of the arrowheads flanking two lines makes them appear as having different lengths), and reported evidence in favour of an absence of correlation between the two measures. This finding was interpreted as indicative of the cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. The goal of this study is thus to replicate the results from @lush2022 pointing to an absence of a relationship between phenomenological control and illusion sensitivity, by generalising them to a different illusion task that would encompass other illusion types.

>>>>

  # - name: Zoltan Dienes
  #   orcid: 0000-0001-7454-3161
  #   roles:
  #     - Supervision
  #     - Writing – review & editing
  #   affiliations: 
  #     - ref: id1

@@

Visual illusions highlight how easily our conscious experience can be altered with respect to perceptual reality. Despite sharing in-principle mechanisms with phenomenological control, i.e., the ability to alter our perceptual experience to match task demands or expectations, research tying the two remains scarce. This study aims to replicate and expand Lush et al. (2022) reporting an absence of correlation between phenomenological control (measured using the Phenomenological Control Scale) and illusion sensitivity to different illusion types. *[N participants were recruited in an online study. Results will be added in the final version of the manuscript]*.

@@

  references-meta-analysis:

@@

Interestingly, @lush2022 did test for a relationship between PC and illusion sensitivity using the Müller-Lyer illusion (in which the arrangement of the arrowheads flanking two lines makes them appear as having different lengths), and reported evidence in favour of an absence of correlation between the two measures. This finding was interpreted as indicative of the cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. The goal of this study is thus to replicate the results from @lush2022 pointing to an absence of a relationship between phenomenological control and illusion sensitivity, by generalising them to a different illusion task that would encompass other illusion types (see table 1).

```{r}
# Create the dataframe
df <- data.frame(
  Question = "Is there a relationship between trait phenomenological control (PC) and visual illusion (VI) sensitivity?",
  Hypothesis = "Replicating findings from Lush et al., 2022 paper, we expect evidence of favour of an absence of relationship between VI and PCS",
  Sampling_plan = "The goal is to recruit around 500 adult English speakers using Prolific. This sample size is based on the ones used in Lush et l., 2021 and Lush et al., 2022 that we aim at replicate.",
  Analysis_Plan = "Bayesian correlations between the PC score and the VI performance for the 3 illusion types (corresponding to the error rate) will be computed using the BayesFactor::BFCorrelation() function (with the r-scale prior parameter set to 'medium')",
  Rationale = "Evidence against a relationship between PC and VI will be found if BF10 <= 1/3, following the Lush et al., 2022 findings. BF10 > 3 will be taken as evidence for a relationship between these two measures.",
  Interpretation = "The hypothesis that VI sensitivity is independent from PC.",
  Theory = ""
)

# Print the dataframe as a table
print(kable(df), format = "markdown")

```

Table 1. Study Design Table
